---
title: "STA6106 Project 1"
author: "Travis Loebs"
date: "October 2, 2016"
output: html_document
sansfont: Times New Roman
fontsize: 12pt
---

Problem 1

1.1 The problem

$$\max_{x_1, x_2, x_3} 20 x_1 + 16 x_2 - 2 x_1^2 - x_2^2 - x_3^2,$$
 
$$s.t. \qquad
x_1 + x_2 \leq 5$$
$$x_1 + x_2 - x_3 = 0$$
$$x_1 \geq 0, x_2 \geq 0, x_3 \geq 0$$

can be expressed in the matrix form 

$$\min_{x_1, x_2, x_3} \frac{1}{2} x' K x + C'x \qquad s.t. \quad Ax + d \leq 0$$

where 

$$x = 
\begin{bmatrix}
x_1 \\
x_2 \\
x_3
\end{bmatrix},
\;
K = 2 \times
\begin{bmatrix}
-2 & 0 & 0 \\
0 & -1 & 0 \\
0 & 0 & -1
\end{bmatrix},
\;
C = 
\begin{bmatrix}
20 & 16 & 0
\end{bmatrix},
\;
A = 
\begin{bmatrix}
1 & 1 & 0 \\
1 & 1 & -1 \\
-1 & -1 & 1 \\
-1 & 0 & 0 \\
0 & -1 & 0 \\
0 & 0 & -1
\end{bmatrix},
\;
d = 
\begin{bmatrix}
-5 \\
0 \\
0 \\
0 \\
0 \\
0 \\
\end{bmatrix}
$$

1.2 This problem can be solved in R using the quadprog library. However, because the quadprog library solves functions of the form

$$\min_b \frac{1}{2} b' D b - d'b \qquad s.t. \quad A'b \geq b_0$$

we must turn our maximization problem into a minimization problem by multiplying the objective and constraint functions by -1. First, we load the library and create our matrices at outlined in problem 1.1, except all of the matrices are multiplied by -1.

```{r comment=NA, message=FALSE, warning=FALSE}
library(quadprog)
K = 2*diag (c (2, 1, 1));
C = c(20, 16, 0);
A = matrix (0, nrow=5, ncol=3);
A[1,] = c(-1, -1, 1);
A[2,] = c(-1, -1, 0);
A[3,] = c(1, 0, 0);
A[4,] = c(0, 1, 0);
A[5,] = c(0, 0, 1);
d = c(0, -5, 0, 0, 0)
```

Using the 'solve.QP' function in R, we can solve this problem and set the estimator, $\hat{x}$, equal to the solution.

```{r comment=NA, message=FALSE, warning=FALSE}
xHat = solve.QP(K, C, t(A), d, meq=1)$solution
xHat
```

We see that this function, subject to the restraints, is maximized at $x_1 = 2 \frac{1}{3}, x_2 = 2 \frac{2}{3},$ and $x_3 = 5$.

&nbsp;

Problem 2

2.1 Given the objective and constraint functions, the Lagrangian is

$$\mathcal{L}(x, \lambda) = x_1^2 + 2 x_2^2 + 2 x_1 + 8 x_2 
+ \lambda_1 [-x_1 - 2 x_2 + 10] - \lambda_2 x_1 - \lambda_3 x_2$$

2.2 To find the dual problem, we first differentiate the Lagrangian with respect to $x$

$$\nabla_x \mathcal{L}(x, \lambda) = \langle
2 x_1 + 2 - \lambda_1 - \lambda_2,
\; 4 x_2 + 8 - 2 \lambda_1 - \lambda_3 \rangle$$

Setting to zero and solving for $x$, we get

$$x_1 = {\frac{\lambda_1 + \lambda_2 - 2}{2}} \; , \quad
x_2 = {\frac{2 \lambda_1 + \lambda_3 - 8}{4}}$$

So, the dual problem becomes

$$\max_{\lambda \geq 0} g(\lambda) = \max_{\lambda}\big[\min_x \mathcal{L}(x, \lambda)\big] = {\Big(\frac{\lambda_1 + \lambda_2 - 2}{2}\Big)}^2 + 2 {\Big( \frac{2 \lambda_1 + \lambda_3 - 8}{4}\Big)}^2 + 5 \lambda_1 + \lambda_2 + 2 \lambda_3 - 18
$$
$$+ \lambda_1 \Big[ -{\frac{\lambda_1 + \lambda_2 - 2}{2}} - 2 {\frac{2 \lambda_1 + \lambda_3 - 8}{4}} + 10 \Big] - \lambda_2 {\frac{\lambda_1 + \lambda_2 - 2}{2}} - \lambda_3 {\frac{2 \lambda_1 + \lambda_3 - 8}{4}}$$
$$= 15 \lambda_1 + \lambda_2 + 2 \lambda_3 - \frac{3}{4}\lambda_1^2 - \frac{1}{2}\lambda_1 \lambda_2 - \frac{1}{2}\lambda_1 \lambda_3 - \frac{1}{2}\lambda_2^2 - \frac{1}{8}\lambda_3^2 - 9$$

2.3 The Karush-Kuhn-Tucker (KKT) conditions are given by:
$$ \lambda_1 \geq 0, \; \lambda_2 \geq 0, \; \lambda_3 \geq 0$$
$$ -x_1 - 2 x_2 + 10 \leq 0$$
$$ -x_1 \leq 0, \; -x_2 \leq 0$$
$$\lambda_1 [-x_1 - 2 x_2 + 10] = 0$$
$$- \lambda_2 x_1 = 0$$
$$- \lambda_3 x_2 = 0$$

2.4 
```{r}
library(quadprog)
K = 2*diag (c (-1, -2));
C = c(2, 8)
A = matrix (0, nrow=3, ncol=2)
A[1,] = c(1, 2)
A[2,] = c(-1, 0)
A[3,] = c(0, -1)
d = c(-10, 0, 0)
#xHat = solve.QP(K, C, t(A), d, meq=0)$solution
#xHat
```

```{r}
library(quadprog)
K = matrix(0, nrow=3, ncol=3)
K[1,] = -2*c(3/4, -1/4, -1/4)
K[2,] = -2*c(-1/4, -1/2, 0)
K[3,] = -2*c(-1/4, 0, -1/8)
C = c(15, 1, 2)
A = matrix (0, nrow=3, ncol=3)
A[1,] = c(1, 0, 0)
A[2,] = c(0, 1, 0)
A[3,] = c(0, 0, 1)
d = c(0, 0, 0)
#xHat = solve.QP(K, C, t(A), d, meq=0)$solution
#xHat
```

